---
title: "Raman single-cell spectroscopy: reproducible analysis"
author: "Jay T. Lennon"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
geometry: margin=2.54cm
header-includes:
  - \usepackage{array}
---

# Overview

This analysis asks whether single-cell Raman spectra distinguish Bacillus endospores from vegetative cells and whether fixation changes Raman signatures.

The primary biological experiment is a 2 x 2 design:

| Cell state | Unfixed | Fixed |
|---|---|---|
| Endospore | psu | psf |
| Vegetative | pvu | pvf |

Soil blanks (sbu/sbf) are retained in the imported dataset but excluded from the biological 2 x 2 analysis.

The processing sequence used here is:

**import -> metadata -> QC for near-ceiling spectra -> CaDPA benchmark -> fingerprint selection -> ALS baseline correction -> L2/vector normalization -> PCA -> acquisition/run QC**.

The CaDPA analysis is intentionally kept separate from the whole-spectrum analysis so that the known endospore-associated feature provides a simple biological benchmark for any later multivariate classifier.

## 1. Setup

```{r setup}
rm(list = ls())

suppressPackageStartupMessages({
  library(here)
  library(data.table)
  library(dplyr)
  library(Matrix)
  library(pROC)
})

# ------------------------------------------------------------
# Project paths
# ------------------------------------------------------------

# here() should resolve to the root of the "raman" repository
project_dir <- here()

# Input data
data_dir <- here("2.data")

# Analysis outputs and figures
output_dir <- here("3.output")

# Raman matrix
input_file <- here(
  "2.data",
  "raman_cell_matrix.tsv"
)


# Consistent colors for the four biological groups
# (used only for figures; not part of the analysis itself)
group_colors <- c(
  "vegetative_unfixed" = "steelblue",
  "endospore_unfixed" = "firebrick",
  "vegetative_fixed" = "skyblue3",
  "endospore_fixed" = "orange3"
)

```

# 2. Import the assembled Raman matrix

The assembled matrix has Raman positions in rows and individual spectra/cells in columns. Column names retain the original processed-file identifiers so that acquisition time can later be recovered from the filename.

```{r import}
raman <- fread(
  input_file,
  sep = "\t",
  header = TRUE,
  check.names = FALSE
)

raman_shift <- as.numeric(raman[[1]])

spectra <- as.matrix(
  raman[, -1, with = FALSE]
)
storage.mode(spectra) <- "numeric"

cell_id <- colnames(spectra)

cat("Raman matrix dimensions:\n")
print(dim(raman))

cat("\nRaman positions:", length(raman_shift), "\n")
cat("Spectra:", ncol(spectra), "\n")
cat(
  "Spectral range:",
  min(raman_shift), "to", max(raman_shift),
  "cm^-1\n"
)
```

# 3. Construct experimental metadata

Numeric prefixes such as `2-`, `3-`, etc. are ordering identifiers, not treatment identities. The final `1` in the sample code denotes replicate 1 in this dataset.

```{r metadata}
# ------------------------------------------------------------
# Sample identity
# ------------------------------------------------------------

sample_id <- sub(
  "^([0-9]+-)?(psu|sbu|psf|sbf|pvf|pvu)1__.*$",
  "\\2",
  cell_id
)

sample_id <- factor(
  sample_id,
  levels = c("psu", "psf", "pvu", "pvf", "sbu", "sbf")
)

# ------------------------------------------------------------
# Biological state and fixation
# ------------------------------------------------------------

metadata <- data.frame(
  cell_id = cell_id,
  sample = sample_id,
  stringsAsFactors = FALSE
)

metadata$cell_state <- case_when(
  metadata$sample %in% c("psu", "psf") ~ "endospore",
  metadata$sample %in% c("pvu", "pvf") ~ "vegetative",
  TRUE ~ NA_character_
)

metadata$fixation <- case_when(
  metadata$sample %in% c("psf", "pvf", "sbf") ~ "fixed",
  metadata$sample %in% c("psu", "pvu", "sbu") ~ "unfixed",
  TRUE ~ NA_character_
)

metadata$replicate <- 1L

metadata$treatment <- case_when(
  metadata$sample == "psu" ~ "pure spores, unfixed",
  metadata$sample == "psf" ~ "pure spores, fixed",
  metadata$sample == "pvu" ~ "pure vegetative cells, unfixed",
  metadata$sample == "pvf" ~ "pure vegetative cells, fixed",
  metadata$sample == "sbu" ~ "soil blank supernatant, unfixed",
  metadata$sample == "sbf" ~ "soil blank supernatant, fixed",
  TRUE ~ NA_character_
)

metadata$sample <- factor(
  metadata$sample,
  levels = c("psu", "psf", "pvu", "pvf", "sbu", "sbf")
)

metadata$cell_state <- factor(
  metadata$cell_state,
  levels = c("vegetative", "endospore")
)

metadata$fixation <- factor(
  metadata$fixation,
  levels = c("unfixed", "fixed")
)

# ------------------------------------------------------------
# Full-data checks
# ------------------------------------------------------------

cat("Cells per sample:\n")
print(table(metadata$sample, useNA = "ifany"))

stopifnot(
  length(raman_shift) == nrow(spectra),
  ncol(spectra) == nrow(metadata),
  identical(colnames(spectra), metadata$cell_id),
  all(!is.na(metadata$sample)),
  all(!is.na(metadata$fixation))
)

cat("\nMetadata validation passed.\n")
```

# 4. Define the biological 2 x 2 experiment

Soil blanks are controls, not biological cells for the present question, and therefore are excluded from the 2 x 2 analysis.

```{r biological_subset}
biological_cells <- !is.na(metadata$cell_state)

spectra_bio <- spectra[, biological_cells, drop = FALSE]
metadata_bio <- droplevels(metadata[biological_cells, ])

metadata_bio$group <- interaction(
  metadata_bio$cell_state,
  metadata_bio$fixation,
  sep = "_",
  drop = TRUE
)

cat("Biological spectra:", ncol(spectra_bio), "\n")
cat("Raman positions:", nrow(spectra_bio), "\n")

cat("\n2 x 2 experimental design:\n")
print(table(metadata_bio$cell_state, metadata_bio$fixation))

stopifnot(
  ncol(spectra_bio) == nrow(metadata_bio),
  identical(colnames(spectra_bio), metadata_bio$cell_id),
  all(!is.na(metadata_bio$cell_state)),
  all(!is.na(metadata_bio$fixation))
)
```

# 5. Initial spectral QC: identify near-ceiling spectra

The processed spectra contain a subset with very large contiguous regions near the upper end of the intensity scale. We used an empirical threshold of **63,000 intensity units** as a QC flag.

This is a data-derived QC threshold, not a claim about the instrument's official detector ceiling. The collaborator memo therefore asks the instrument team to confirm what the processed intensity scale and true saturation limit are.

```{r saturation_qc}
# Number of Raman positions at or above the empirical near-ceiling threshold
n_high <- colSums(spectra_bio >= 63000)

spectrum_max <- apply(
  spectra_bio,
  2,
  max,
  na.rm = TRUE
)

qc <- data.frame(
  cell_id = metadata_bio$cell_id,
  sample = metadata_bio$sample,
  cell_state = metadata_bio$cell_state,
  fixation = metadata_bio$fixation,
  spectrum_max = spectrum_max,
  n_high = n_high
)

cat("\nSpectra with at least one value >= 63,000:\n")
print(table(qc$n_high > 0))

cat("\nDistribution of number of near-ceiling points:\n")
print(quantile(
  qc$n_high,
  probs = c(0, .5, .9, .95, .99, .995, .999, 1)
))

cat("\nQC counts by biological group:\n")
print(
  qc %>%
    mutate(near_ceiling = n_high > 0) %>%
    count(cell_state, fixation, near_ceiling)
)

# Keep spectra without any near-ceiling point
qc_pass <- qc$n_high == 0

metadata_bio$n_high <- qc$n_high
metadata_bio$spectrum_max <- qc$spectrum_max
metadata_bio$qc_near_ceiling <- !qc_pass

spectra_qc <- spectra_bio[, qc_pass, drop = FALSE]
metadata_qc <- droplevels(metadata_bio[qc_pass, ])

stopifnot(
  ncol(spectra_qc) == nrow(metadata_qc),
  identical(colnames(spectra_qc), metadata_qc$cell_id),
  all(is.finite(spectra_qc))
)

cat("\nSpectra before QC:", ncol(spectra_bio), "\n")
cat("Near-ceiling spectra removed:", sum(!qc_pass), "\n")
cat("Spectra after QC:", ncol(spectra_qc), "\n")

cat("\nFinal 2 x 2 design after QC:\n")
print(table(metadata_qc$cell_state, metadata_qc$fixation))
```

# 6. CaDPA-associated Raman feature: descriptive benchmark

A Raman position near **1016 cm^-1** is retained as a simple, biologically motivated benchmark for endospore classification.

The measured Raman position nearest 1016 cm^-1 is identified directly from the imported axis rather than hard-coding a row number.

## 6.1 Local baseline correction at the CaDPA-associated feature

```{r cdpa_metric}
# Measured Raman position nearest 1016 cm^-1
peak_index <- which.min(abs(raman_shift - 1016))
peak_shift <- raman_shift[peak_index]

cat("Raman position nearest 1016 cm^-1:", peak_shift, "\n")

# Local baseline anchors on either side of the peak
left_region <- raman_shift >= 980 & raman_shift <= 995
right_region <- raman_shift >= 1035 & raman_shift <= 1050

left_intensity <- colMeans(
  spectra_qc[left_region, , drop = FALSE]
)

right_intensity <- colMeans(
  spectra_qc[right_region, , drop = FALSE]
)

left_shift <- mean(raman_shift[left_region])
right_shift <- mean(raman_shift[right_region])

# Linear interpolation between the two baseline regions
baseline_at_peak <- left_intensity +
  (right_intensity - left_intensity) *
  ((peak_shift - left_shift) / (right_shift - left_shift))

peak_intensity <- spectra_qc[peak_index, ]

cdpa_peak <- peak_intensity - baseline_at_peak

metadata_qc$cdpa_peak <- cdpa_peak

cdpa_stats <- metadata_qc %>%
  group_by(cell_state, fixation) %>%
  summarise(
    n = n(),
    mean = mean(cdpa_peak),
    median = median(cdpa_peak),
    sd = sd(cdpa_peak),
    q25 = quantile(cdpa_peak, .25),
    q75 = quantile(cdpa_peak, .75),
    p05 = quantile(cdpa_peak, .05),
    p95 = quantile(cdpa_peak, .95),
    .groups = "drop"
  )

print(cdpa_stats)
```

## 6.2 Visualize the benchmark

```{r cdpa_plot, fig.width=7, fig.height=5}
pdf(
  file = file.path(output_dir, "Fig.1.pdf"),
  width = 7,
  height = 5
)

boxplot(
  cdpa_peak ~ cell_state * fixation,
  data = metadata_qc,
  names = c(
    "Vegetative\nUnfixed",
    "Endospore\nUnfixed",
    "Vegetative\nFixed",
    "Endospore\nFixed"
  ),
  xlab = "",
  ylab = expression("Baseline-corrected intensity near " * cm^{-1}),
  main = "CaDPA-associated Raman peak"
)

abline(
  h = 0,
  lty = 2,
  col = "gray60"
)

dev.off()
```

## 6.3 ROC/AUC benchmark

The binary outcome is vegetative = 0 and endospore = 1. The ROC analysis is a **descriptive benchmark** for this replicate, not a claim of externally validated predictive performance.

```{r cdpa_roc}
metadata_qc$is_endospore <- ifelse(
  metadata_qc$cell_state == "endospore",
  1,
  0
)

roc_all <- roc(
  response = metadata_qc$is_endospore,
  predictor = metadata_qc$cdpa_peak,
  levels = c(0, 1),
  direction = "<",
  quiet = TRUE
)

unfixed <- metadata_qc$fixation == "unfixed"
fixed <- metadata_qc$fixation == "fixed"

roc_unfixed <- roc(
  response = metadata_qc$is_endospore[unfixed],
  predictor = metadata_qc$cdpa_peak[unfixed],
  levels = c(0, 1),
  direction = "<",
  quiet = TRUE
)

roc_fixed <- roc(
  response = metadata_qc$is_endospore[fixed],
  predictor = metadata_qc$cdpa_peak[fixed],
  levels = c(0, 1),
  direction = "<",
  quiet = TRUE
)

cat("All cells AUC:", round(as.numeric(auc(roc_all)), 3), "\n")
cat("Unfixed AUC:", round(as.numeric(auc(roc_unfixed)), 3), "\n")
cat("Fixed AUC:", round(as.numeric(auc(roc_fixed)), 3), "\n")

cat("\n95% CI for AUC:\n")
print(ci.auc(roc_all))
print(ci.auc(roc_unfixed))
print(ci.auc(roc_fixed))

cat("\nUnfixed vs fixed ROC comparison:\n")
print(roc.test(roc_unfixed, roc_fixed, paired = FALSE))
```

```{r cdpa_roc_plot, fig.width=7, fig.height=5}
roc_all_x <- 1 - roc_all$specificities
roc_all_y <- roc_all$sensitivities

roc_unfixed_x <- 1 - roc_unfixed$specificities
roc_unfixed_y <- roc_unfixed$sensitivities

roc_fixed_x <- 1 - roc_fixed$specificities
roc_fixed_y <- roc_fixed$sensitivities

# ------------------------------------------------------------
# Figure 2. ROC curves for CaDPA-based classification
# ------------------------------------------------------------

plot_fig2 <- function() {

  plot(
    roc_all_x,
    roc_all_y,
    type = "l",
    lwd = 2,
    xlim = c(0, 1),
    ylim = c(0, 1),
    xaxs = "i",
    yaxs = "i",
    xlab = "1 - Specificity",
    ylab = "Sensitivity",
    main = "Endospore classification using CaDPA Raman peak"
  )

  lines(
    roc_unfixed_x,
    roc_unfixed_y,
    lwd = 2,
    lty = 2
  )

  lines(
    roc_fixed_x,
    roc_fixed_y,
    lwd = 2,
    lty = 3
  )

  abline(
    a = 0,
    b = 1,
    lty = 3,
    col = "gray60"
  )

  legend(
    "bottomright",
    legend = c(
      paste0(
        "All cells (AUC = ",
        round(as.numeric(auc(roc_all)), 3),
        ")"
      ),
      paste0(
        "Unfixed (AUC = ",
        round(as.numeric(auc(roc_unfixed)), 3),
        ")"
      ),
      paste0(
        "Fixed (AUC = ",
        round(as.numeric(auc(roc_fixed)), 3),
        ")"
      )
    ),
    lty = c(1, 2, 3),
    lwd = 2,
    bty = "n"
  )
}


# Display Figure 2 in R Markdown
plot_fig2()


# Save Figure 2 as a PDF in 3.output
pdf(
  file = file.path(output_dir, "Fig.2.pdf"),
  width = 7,
  height = 5
)

plot_fig2()

dev.off()
```

# 7. Whole-spectrum preprocessing: fingerprint region

The main multivariate analysis uses the **600-1800 cm^-1 fingerprint region**. We do not use the full 336-4279 cm^-1 range here because the lower fingerprint region contains the strongest interpretable biochemical structure in the present analysis and limits the influence of broad high-wavenumber features.

```{r fingerprint}
fingerprint <- raman_shift >= 600 & raman_shift <= 1800

raman_fingerprint <- raman_shift[fingerprint]
spectra_fp <- spectra_qc[fingerprint, , drop = FALSE]

cat("Fingerprint matrix dimensions:\n")
print(dim(spectra_fp))
```

# 8. ALS baseline correction

Baseline correction was necessary because L2-normalized spectra before baseline correction produced a PCA in which PC1 explained about 92% of total variance and was dominated by broad spectral/background structure.

The working ALS parameters selected during the pilot were:

- `lambda = 1e5`
- `p = 0.01`
- `niter = 10`

The pilot was checked on endospore and vegetative spectra before being applied to all QC-passing cells.

```{r als_function}
# ------------------------------------------------------------
# Asymmetric least-squares (ALS) baseline estimator
# ------------------------------------------------------------

als_baseline <- function(
  y,
  lambda = 1e5,
  p = 0.01,
  niter = 10
) {

  n <- length(y)

  # Second-difference penalty matrix
  D <- diff(Diagonal(n), differences = 2)

  w <- rep(1, n)

  for (i in seq_len(niter)) {

    W <- Diagonal(x = w)

    z <- solve(
      W + lambda * crossprod(D),
      w * y
    )

    z <- as.numeric(z)

    # Down-weight points above the estimated baseline
    w <- ifelse(y > z, p, 1 - p)
  }

  z
}
```

## 8.1 Optional parameter check

This small diagnostic can be rerun if a future dataset differs materially from this one. It is deliberately kept separate from the main pipeline because the present dataset already provided a working parameter choice.

```{r als_pilot, eval=FALSE}
set.seed(123)

pilot_cells <- unlist(
  lapply(
    split(seq_len(ncol(spectra_qc)), metadata_qc$group),
    function(x) sample(x, min(5, length(x)))
  )
)

pilot_groups <- metadata_qc$group[pilot_cells]
pilot_raw <- t(spectra_fp[, pilot_cells, drop = FALSE])

example <- which(pilot_groups == "endospore_unfixed")[1]
raw_example <- pilot_raw[example, ]

baseline_1e4 <- als_baseline(raw_example, lambda = 1e4, p = 0.01)
baseline_1e5 <- als_baseline(raw_example, lambda = 1e5, p = 0.01)
baseline_1e6 <- als_baseline(raw_example, lambda = 1e6, p = 0.01)

# ------------------------------------------------------------
# Figure 3. ALS baseline comparison
# ------------------------------------------------------------

plot_fig3 <- function() {

  plot(
    raman_fingerprint,
    raw_example,
    type = "l",
    lwd = 1.5,
    xlab = expression("Raman shift (" * cm^{-1} * ")"),
    ylab = "Intensity",
    main = "ALS baseline comparison"
  )

  lines(
    raman_fingerprint,
    baseline_1e4,
    lty = 2,
    lwd = 2
  )

  lines(
    raman_fingerprint,
    baseline_1e5,
    lty = 3,
    lwd = 2
  )

  lines(
    raman_fingerprint,
    baseline_1e6,
    lty = 4,
    lwd = 2
  )

  legend(
    "topright",
    legend = c(
      "Raw",
      "lambda = 1e4",
      "lambda = 1e5",
      "lambda = 1e6"
    ),
    lty = c(1, 2, 3, 4),
    lwd = c(1.5, 2, 2, 2),
    bty = "n"
  )
}


# Display Figure 3 in R Markdown
plot_fig3()


# Save Figure 3 as a PDF in 3.output
pdf(
  file = file.path(output_dir, "Fig.3.pdf"),
  width = 7,
  height = 5
)

plot_fig3()

dev.off()
```

## 8.2 Apply ALS correction to all QC-passing spectra

```{r als_full}
cat("Starting ALS baseline correction...\n")

spectra_bc <- apply(
  spectra_fp,
  2,
  function(y) {
    y - als_baseline(
      y,
      lambda = 1e5,
      p = 0.01,
      niter = 10
    )
  }
)

colnames(spectra_bc) <- colnames(spectra_fp)

stopifnot(
  ncol(spectra_bc) == nrow(metadata_qc),
  identical(colnames(spectra_bc), metadata_qc$cell_id),
  all(is.finite(spectra_bc))
)

cat("Baseline correction complete.\n")
print(dim(spectra_bc))
```

# 9. L2/vector normalization after baseline correction

Each cell spectrum is normalized independently by its Euclidean/L2 norm. This removes overall intensity scale while preserving the shape of each corrected spectrum.

The order is important:

**baseline correction -> L2 normalization**

rather than normalizing first and then estimating the baseline.

```{r l2_normalization}
bc_norm <- sqrt(colSums(spectra_bc^2))

stopifnot(all(bc_norm > 0))

spectra_bc_norm <- sweep(
  spectra_bc,
  2,
  bc_norm,
  "/"
)

norm_check <- sqrt(colSums(spectra_bc_norm^2))

cat("Normalized spectrum lengths:\n")
print(summary(norm_check))

stopifnot(all(abs(norm_check - 1) < 1e-10))

cat("\nL2 normalization validated.\n")
```

# 10. Save the reproducible processing checkpoint

Save processed objects to 3.output so that the computationally expensive baseline-correction step does not need to be repeated after restarting R.

```{r save_checkpoint}
saveRDS(
  spectra_bc_norm,
  file.path(output_dir, "spectra_bc_norm.rds")
)

saveRDS(
  spectra_bc,
  file.path(output_dir, "spectra_bc.rds")
)

saveRDS(
  metadata_qc,
  file.path(output_dir, "metadata_qc.rds")
)

saveRDS(
  raman_fingerprint,
  file.path(output_dir, "raman_fingerprint.rds")
)

cat("Processed Raman objects saved to:\n", output_dir, "\n")
```

# 11. Median spectra after baseline correction and normalization

These medians provide an interpretable overview before moving to multivariate analysis.

```{r median_spectra, fig.width=8, fig.height=5}
metadata_qc$group <- interaction(
  metadata_qc$cell_state,
  metadata_qc$fixation,
  sep = "_",
  drop = TRUE
)

group_levels <- levels(metadata_qc$group)

median_spectra_final <- sapply(
  group_levels,
  function(g) {
    apply(
      spectra_bc_norm[, metadata_qc$group == g, drop = FALSE],
      1,
      median
    )
  }
)

colnames(median_spectra_final) <- group_levels

plot_colors <- group_colors[colnames(median_spectra_final)]

# ------------------------------------------------------------
# Figure 4. Median Raman spectra after preprocessing
# ------------------------------------------------------------

plot_fig4 <- function() {

  matplot(
    raman_fingerprint,
    median_spectra_final,
    type = "l",
    lty = 1,
    lwd = 1.5,
    col = plot_colors,
    xlab = expression("Raman shift (" * cm^{-1} * ")"),
    ylab = "Normalized, baseline-corrected intensity",
    main = "Median Raman spectra after preprocessing"
  )

  # Mark the CaDPA-associated Raman position
  abline(
    v = peak_shift,
    lty = 2,
    col = "gray60"
  )

  legend(
    "topright",
    legend = colnames(median_spectra_final),
    col = plot_colors,
    lty = 1,
    lwd = 2,
    bty = "n"
  )
}


# Display Figure 4 in R Markdown
plot_fig4()


# Save Figure 4 as a PDF in 3.output
pdf(
  file = file.path(output_dir, "Fig.4.pdf"),
  width = 7,
  height = 5
)

plot_fig4()

dev.off()
```

# 12. PCA of the corrected fingerprint spectra

PCA is used here for **exploration and dimension reduction**, not as the final classifier.

Rows in the PCA matrix are cells; columns are Raman positions.

```{r pca}
pca_input <- t(spectra_bc_norm)

pca_final <- prcomp(
  pca_input,
  center = TRUE,
  scale. = FALSE
)

variance_final <- 100 *
  pca_final$sdev^2 /
  sum(pca_final$sdev^2)

cat("Variance explained by first 10 PCs:\n")
print(round(variance_final[1:10], 2))

pca_scores_final <- data.frame(
  PC1 = pca_final$x[, 1],
  PC2 = pca_final$x[, 2],
  PC3 = pca_final$x[, 3],
  PC4 = pca_final$x[, 4],
  metadata_qc,
  row.names = NULL
)

point_colors <- group_colors[
  as.character(metadata_qc$group)
]
```

## 12.1 PCA score plots

```{r pca_scores, fig.width=8, fig.height=6}
plot(
  pca_scores_final$PC1,
  pca_scores_final$PC2,
  pch = 16,
  cex = 0.5,
  col = point_colors,
  xlab = paste0("PC1 (", round(variance_final[1], 1), "%)"),
  ylab = paste0("PC2 (", round(variance_final[2], 1), "%)"),
  main = "PCA after baseline correction and normalization"
)

legend(
  "topright",
  legend = names(group_colors),
  col = group_colors,
  pch = 16,
  bty = "n"
)
```

```{r pca_scores_2, fig.width=8, fig.height=6}
# ------------------------------------------------------------
# Figure 5. PCA of preprocessed Raman spectra
# ------------------------------------------------------------

plot_fig5 <- function() {

  # Two panels side-by-side
  par(mfrow = c(1, 2))

  # ----------------------------------------------------------
  # Panel A: PC1 vs PC2
  # ----------------------------------------------------------

  plot(
    pca_scores_final$PC1,
    pca_scores_final$PC2,
    pch = 16,
    cex = 0.5,
    col = point_colors,
    xlab = paste0(
      "PC1 (",
      round(variance_final[1], 1),
      "%)"
    ),
    ylab = paste0(
      "PC2 (",
      round(variance_final[2], 1),
      "%)"
    ),
    main = "A"
  )

  legend(
    "topright",
    legend = names(group_colors),
    col = group_colors,
    pch = 16,
    bty = "n",
    cex = 0.8
  )


  # ----------------------------------------------------------
  # Panel B: PC1 vs PC3
  # ----------------------------------------------------------

  plot(
    pca_scores_final$PC1,
    pca_scores_final$PC3,
    pch = 16,
    cex = 0.5,
    col = point_colors,
    xlab = paste0(
      "PC1 (",
      round(variance_final[1], 1),
      "%)"
    ),
    ylab = paste0(
      "PC3 (",
      round(variance_final[3], 1),
      "%)"
    ),
    main = "B"
  )

  legend(
    "topright",
    legend = names(group_colors),
    col = group_colors,
    pch = 16,
    bty = "n",
    cex = 0.8
  )

  # Reset plotting layout
  par(mfrow = c(1, 1))
}


# Display Figure 5 in R Markdown
plot_fig5()


# Save Figure 5 as a PDF in 3.output
pdf(
  file = file.path(output_dir, "Fig.5.pdf"),
  width = 12,
  height = 5.5
)

plot_fig5()

dev.off()
```

## 12.2 PCA loadings

```{r pca_loadings, fig.width=8, fig.height=5}
loadings_final <- data.frame(
  raman_shift = raman_fingerprint,
  PC1 = pca_final$rotation[, 1],
  PC2 = pca_final$rotation[, 2],
  PC3 = pca_final$rotation[, 3]
)

# ------------------------------------------------------------
# Figure 6. PCA loadings after preprocessing
# ------------------------------------------------------------

plot_fig6 <- function() {

  # Three panels side-by-side
  par(mfrow = c(1, 3))

  # ----------------------------------------------------------
  # Panel A: PC1 loadings
  # ----------------------------------------------------------

  plot(
    loadings_final$raman_shift,
    loadings_final$PC1,
    type = "l",
    lwd = 1.5,
    xlab = expression("Raman shift (" * cm^{-1} * ")"),
    ylab = "Loading",
    main = "A"
  )

  abline(
    h = 0,
    lty = 2,
    col = "gray60"
  )

  abline(
    v = peak_shift,
    lty = 2,
    col = "gray60"
  )


  # ----------------------------------------------------------
  # Panel B: PC2 loadings
  # ----------------------------------------------------------

  plot(
    loadings_final$raman_shift,
    loadings_final$PC2,
    type = "l",
    lwd = 1.5,
    xlab = expression("Raman shift (" * cm^{-1} * ")"),
    ylab = "Loading",
    main = "B"
  )

  abline(
    h = 0,
    lty = 2,
    col = "gray60"
  )

  abline(
    v = peak_shift,
    lty = 2,
    col = "gray60"
  )


  # ----------------------------------------------------------
  # Panel C: PC3 loadings
  # ----------------------------------------------------------

  plot(
    loadings_final$raman_shift,
    loadings_final$PC3,
    type = "l",
    lwd = 1.5,
    xlab = expression("Raman shift (" * cm^{-1} * ")"),
    ylab = "Loading",
    main = "C"
  )

  abline(
    h = 0,
    lty = 2,
    col = "gray60"
  )

  abline(
    v = peak_shift,
    lty = 2,
    col = "gray60"
  )

  # Reset plotting layout
  par(mfrow = c(1, 1))
}


# Display Figure 6 in R Markdown
plot_fig6()


# Save Figure 6 as a PDF in 3.output
pdf(
  file = file.path(output_dir, "Fig.6.pdf"),
  width = 15,
  height = 5
)

plot_fig6()

dev.off()
```

# 13. PCA summaries by biological group

The PCA score summaries are descriptive. Because replicate 1 has treatment and acquisition time confounded, they should not be interpreted as independent tests of the treatment effect.

```{r pca_group_summary}
pc_group_summary <- pca_scores_final %>%
  group_by(cell_state, fixation) %>%
  summarise(
    n = n(),
    PC1_mean = mean(PC1),
    PC1_sd = sd(PC1),
    PC2_mean = mean(PC2),
    PC2_sd = sd(PC2),
    PC3_mean = mean(PC3),
    PC3_sd = sd(PC3),
    .groups = "drop"
  )

print(pc_group_summary)
```

# 14. Acquisition-time / run-effect diagnostics

Cell identifiers contain acquisition timestamps. These are extracted and compared with PCA scores to determine whether apparent spectral substructure tracks acquisition history.

This is a **diagnostic** section. We do not regress acquisition time out of the spectra because, in replicate 1, acquisition order is strongly confounded with treatment.

```{r acquisition_time}
# Example cell ID component:
# ...processed20260818-131909.285

time_string <- sub(
  "^.*processed[0-9]{8}-([0-9]{6}\\.[0-9]+)$",
  "\\1",
  metadata_qc$cell_id
)

hh <- as.numeric(substr(time_string, 1, 2))
mm <- as.numeric(substr(time_string, 3, 4))
ss <- as.numeric(substr(time_string, 5, nchar(time_string)))

metadata_qc$acquisition_seconds <- hh * 3600 + mm * 60 + ss

stopifnot(all(!is.na(metadata_qc$acquisition_seconds)))

pca_scores_final$acquisition_seconds <-
  metadata_qc$acquisition_seconds

# Correlation of each of the first three PCs with acquisition time,
# calculated within sample so treatment is not mixed across groups.
time_correlations <- pca_scores_final %>%
  group_by(sample) %>%
  summarise(
    n = n(),
    cor_PC1_time = cor(PC1, acquisition_seconds),
    cor_PC2_time = cor(PC2, acquisition_seconds),
    cor_PC3_time = cor(PC3, acquisition_seconds),
    .groups = "drop"
  )

print(time_correlations)
```

## 14.1 Acquisition order plots

```{r acquisition_order, fig.width=8, fig.height=7}
# ------------------------------------------------------------
# Figure 7. PC3 trajectories across acquisition order
# ------------------------------------------------------------

# Calculate acquisition order within each sample
metadata_qc$acquisition_order <- ave(
  metadata_qc$acquisition_seconds,
  metadata_qc$sample,
  FUN = function(x) rank(x, ties.method = "first")
)

pca_scores_final$acquisition_order <-
  metadata_qc$acquisition_order


# ------------------------------------------------------------
# Plotting function
# ------------------------------------------------------------

plot_fig7 <- function() {

  # Four panels
  par(mfrow = c(2, 2))

  for (s in levels(metadata_qc$sample)) {

    idx <- metadata_qc$sample == s

    plot(
      pca_scores_final$acquisition_order[idx],
      pca_scores_final$PC3[idx],
      pch = 16,
      cex = 0.4,
      xlab = "Acquisition order",
      ylab = "PC3 score",
      main = s
    )
  }

  # Reset plotting layout
  par(mfrow = c(1, 1))
}


# ------------------------------------------------------------
# Display Figure 7 in R Markdown
# ------------------------------------------------------------

plot_fig7()


# ------------------------------------------------------------
# Save Figure 7 as PDF in 3.output
# ------------------------------------------------------------

pdf(
  file = file.path(output_dir, "Fig.7.pdf"),
  width = 8,
  height = 8
)

plot_fig7()

dev.off()
```

## 14.2 Identify unusually long acquisition gaps

A 60-second threshold is used here only as a **diagnostic flag** for conspicuous pauses. It is not treated as a definitive instrument-event definition.

```{r acquisition_gaps}
run_qc <- data.frame(
  sample = metadata_qc$sample,
  cell_id = metadata_qc$cell_id,
  time = metadata_qc$acquisition_seconds,
  PC1 = pca_final$x[, 1],
  PC2 = pca_final$x[, 2],
  PC3 = pca_final$x[, 3]
) %>%
  arrange(sample, time) %>%
  group_by(sample) %>%
  mutate(
    acquisition_order = row_number(),
    gap_seconds = time - lag(time)
  ) %>%
  ungroup()

long_gaps <- run_qc %>%
  filter(gap_seconds >= 60) %>%
  select(
    sample,
    acquisition_order,
    cell_id,
    gap_seconds,
    PC1,
    PC2,
    PC3
  )

print(long_gaps)

cat("\nNumber of acquisition gaps >= 60 seconds by sample:\n")
print(count(long_gaps, sample))
```

# 15. Important interpretation notes

1. **The four biological groups are endospore/vegetative x fixed/unfixed.** Soil blanks are not part of the biological 2 x 2.

2. **Near-ceiling spectra were removed before biological analysis.** The 63,000 threshold is an empirical QC flag and should be reconciled with the instrument team's documented signal range.

3. **The CaDPA feature near 1016.826 cm^-1 is a useful benchmark but not the whole fingerprint.** Whole-spectrum PCA shows additional informative regions.

4. **Baseline correction changes the multivariate representation substantially.** Before correction, PC1 dominated the PCA; after ALS correction and L2 normalization, variance is distributed across multiple PCs.

5. **Acquisition/run structure is real in this dataset.** Some apparent PCA subclusters track acquisition order, including abrupt transitions following long acquisition gaps.

6. **Treatment and acquisition are confounded in replicate 1.** Therefore replicate 1 is suitable for exploration and model development but is not sufficient to establish generalizable treatment effects or independently validated classification performance.

7. **Future validation should be independent of acquisition block and, ideally, independent biological replicate.** A random cell-level train/test split would risk leakage because neighboring spectra can share acquisition-specific structure.

# 16. Current analysis objects

At completion, the main objects are:

- `spectra_qc`: QC-passing spectra in the original full spectral range.
- `spectra_bc`: ALS baseline-corrected spectra in the 600-1800 cm^-1 fingerprint region.
- `spectra_bc_norm`: baseline-corrected + L2-normalized fingerprint spectra.
- `metadata_qc`: metadata aligned exactly to columns of `spectra_qc` and `spectra_bc_norm`.
- `pca_final`: PCA model of the corrected, normalized fingerprint spectra.
- `pca_scores_final`: cell-level PCA scores plus metadata.
- `run_qc`: acquisition-time diagnostics.
